---
title: "PCA on mtcars dataset"
author: "Sahir Rai Bhatnagar"
date: "March 23, 2016"
output:
  html_document:
    fig_height: 7
    fig_width: 9
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

```

In the following document, we perform principal component analysis of the `datasets::mtcars` data

# Introducing the `mtcars` dataset

Here is the actual dataset

```{r, results='asis'}
knitr::kable(mtcars)
```

And here are the summary statistics for each variable

```{r}
summary(mtcars)
```



# Principal Component Analysis on the `mtcars` dataset

```{r}
# cor = TRUE indicates that PCA is performed on 
# standardized data (mean = 0, variance= 1)
pcaCars <- princomp(mtcars, cor = TRUE)

# view objects stored in pcaCars
names(pcaCars)

# proportion of variance explained
summary(pcaCars)
```

## Visulation of the PCs

```{r}
# bar plot
plot(pcaCars)

# scree plot
plot(pcaCars, type = "l")

```


# Cluster Analysis Using PCA Scores

First we cluster the cars using hierarchical clustering

```{r}
# cluster cars
carsHC <- hclust(dist(pcaCars$scores), method = "ward.D2")

# dendrogram
plot(carsHC)

```


## Cutting the Dendrogram

```{r}
# cut the dendrogram into 3 clusters
carsClusters <- cutree(carsHC, k = 3)

# draw dendogram with red borders around the 3 clusters 
plot(carsHC)
rect.hclust(carsHC, k=3, border="red")
```


## First 2 PCs with Cluster Membership

```{r}
# add cluster to data frame of scores
carsDf <- data.frame(pcaCars$scores, "cluster" = factor(carsClusters))
str(carsDf)

# plot the first 2 PCs with cluster membership
# need to install ggplot2 and ggrepel packages first
# using the following command in R: 
# install.packages(c("ggplot2","ggrepel"))
library(ggplot2)
library(ggrepel)
ggplot(carsDf,aes(x=Comp.1, y=Comp.2)) +
  geom_text_repel(aes(label = rownames(carsDf))) +
  theme_classic() +
  geom_hline(yintercept = 0, color = "gray70") +
  geom_vline(xintercept = 0, color = "gray70") +
  geom_point(aes(color = cluster), alpha = 0.55, size = 3) +
  xlab("PC1") +
  ylab("PC2") + 
  xlim(-5, 6) + 
  ggtitle("PCA plot of Cars")
```


